### 2009-12-08 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
First Draft of the dissertation peer reviewed with Kleber.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@680 559edde2-a4d4-44e0-96fc-4035e40d7396

In order to provide information to users, the RTC scientists use OPeNDAP
\cite{opendap} server to distribute collected data over the Internet. As
discussed in the previous chapter, this work proposes a data persistence
application of a Key-Value database system in order to provide an alternative
method for data persistence in sensor networks, taking into account the
taxonomies discussed earlier in chapter \ref{chap:taxonomies}.

Considering our case study being NetBEAMS, this chapter describes the design
and architecture component designed specifically for the Data Sensor Platform,
DSP, in order to provide a persistence layer for NetBEAMS. It is vital to note
that the technology chosen in the last chapter can be adapted by any
alternative that uses the data model and infrastructure options presented.

The proposed DSP Data Persistence component for the DSP
Platform was implemented by using the same approaches described in chapter
\ref{chap:netbeams-overview}. The following sections describe the scenarios for
persistence, analyze the requirements from RTC, and describe the design and
architecture of the solution.

@@

As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco Bay,
composed by different types of sensors. The process of extracting the
measurements from sensors can be summarized following the UML Business Process
Diagram \cite{uml} depicted in figure
\ref{fig:dsp-persistence-business-process}. It is important to note that this
diagram details remote communication between sensor devices and a centralized
server, as part of the main scenario to solve the problem of data persistence.

@@

At the event of data collection, the DSP Client receives measurements collected
from each of the sensor devices. Each of the co-located DSP Client process the
collected measurements and transform them into the alleged DSP Message
(section \ref{sec:dsp-message}), which embodies the actual data values. After
proper data processing, transformation and categorization, the DSP Message is
added into temporary outbound queue. Then, as an asynchronous data transmission
event occurs with a specified rate, transferring all the queued messages to be
processed by a DSP Server host. As image
\ref{fig:dsp-persistence-business-process}, the server is centralized and able
to concurrently process messages.

This work proposes the addition of a database system that is responsible for
storing the gathered measurement data from sensor nodes of the SF-BEAMS through
the use of NetBEAMS. Also in figure \ref{fig:dsp-persistence-business-process},
the processed measurement data from sensors 1 through N must be collected in
the same style and categorized as necessary. In this way, end users such as
Biologists or Programmers can access the collected measurements directly
from the Database. Furthermore, the same collected data can be exported to
different formats as necessary.

As a matter of fact, the data measurements collected from the sensor devices
are basic properties with keys and associated values as described in section 
\ref{sec:sfbeams}, after data has been described. According to the descriptions
of the DSP Platform and the DSP Messages structure, the data is represented in
programming language formats such as Java \cite{java} or in a serialized
version in XML \cite{xml}. Similarly, the collected data contains measurements
that uniquely identifies them, together with other properties such as
timestamps of data generation and collected and the IP address of the DSP
Client that produced the data. Therefore, these are essential properties for
the proposed solution of this work.

@@

By analyzing the main scenario of collected data in the previous section, as
well as by collected requirements from the RTC staff, this section lists the
main use cases that the system and the primary users require.

@@

As part of the user stories related to data access, figure
\ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users} describes the use cases
that the end users can perform with the system.

@@

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: users of the
  collected data must be able to search data through the properties of the
  sensors. This is related to identity principle of what was collected
  as provenance \cite{db-provenance};
  \item \textbf{Search Measurements by Sensor Location}: users of the collected
  data must be able to search data by a given sensor location, as referred to
  the provenance from where the data was collected;
  \item \textbf{Search Measurements by Data collection date}: users of the
  collected data must be able to search data by a given date, or ranges of
  start and end date;
  \item \textbf{Access the data through API, Programming Languages}: users of the
  collected data must be able to search data by the use of different
  programming and scripting languages of their preference;
  \item \textbf{Export Data to OPEnDAP format}: users of the collected data
  must be able to export the collected data to the OPEnDAP format;
  \item \textbf{Export Data to Spreadsheets}: users of the collected data must
  be able to export the collected data to spreadsheets.
\end{itemize}

@@

Different constraints are required to execute the functional requirements
described in the previous section. They are listed as non-functional
requirements as follows:

\begin{itemize}
  \item The DSP Data Persistence component must maintain the received DSP
  Measurement Messages in-memory for a specific period of time in order to
  decrease the write load on the database;
  \item The Persistence storage system must be able to cope with dozens of DSP
  Messages received from different sensors;
  \item The Persistence storage system must be able to save measurements quickly;
  \item The Persistence storage system must be able to scale without too much
  architectural changes, or schema changes;
  \item The Persistence storage system must not impose the use of
  database languages such as SQL \cite{sql} or XML XPath \cite{xml-xpath}
  related to the data being persisted. Instead, data must be handled through API
  calls, as better abstractions for non-computer science researchers
  \cite{sn-programming-language}.
\end{itemize}

@@

As described in the previous chapter, the data model chosen to support
persistence of collected data from sensor networks is the Key-Value-Pair Data
Model \cite{db-kvp}. Beyond the data collected from the sensor devices, Data
Provenance techniques must also be considered to better describe the collected
data \cite{sn-provenance}. In general, the data model must include the
following metadata besed on Data Provenance:

@@

After analyzing the DSP Platform and the structure of the DSP Messages in
chapter \ref{netbeams-architecture}, the following set of keys are expected to
be part of the design of the data model. Any key represented by time or
location are related to the provenance support discussed in section
\ref{sec:sn-data-description}:

\begin{itemize}
  \item \textbf{sensor.ip\underline{ }address}: it is extracted
  from the DSP Message Produce and identifies which sensor generated the sampling;
  \item \textbf{sensor.location.latitude, longitude}: the latitude and
  longitude GPS coordinates of the position of the sensor device, captured from
  references of the device. This is currently provided as a parameter in the
  bootstrap message for the component;
  \item \textbf{message\underline{ }id}: it's extracted from each of the
  messages contained in the DSP Message Container;
  \item \textbf{time.transaction}: it is extracted from the DSP
  message container creation time and it is used to identify when the
  transaction occurred;
  \item \textbf{time.fact}: it is extracted from the
  SondeDataContainer's date and time, and identifies the time in which the
  collected data occurred.
\end{itemize}

It is important to note that this set of metadata must occur in every
collected data from any sensor device. However, on the contrary of the
metadata, a different set of properties will occur to describe the sensed
properties of a given sensor, and for this reason, the following structure is
defined to accommodate such data:

\begin{itemize}
  \item \textbf{observation}: this key defines the values of the document, and
  will have every different property of the sensor. For instance, the data key
  for the properties of the YSI device will contain the set of properties 
  defined for the device, such as battery, temperature, salinity, etc.
\end{itemize}

@@

Considering the business analysis described in section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding Data Consumer (DC) as described
in section \ref{netbeams-architecture}. Based on the Requirements
specification described earlier, this component must be responsible for
providing the processing of the DSP Measure Messages, extracting the payload of
the body of the message and converting it to the data model defined in the
previous section. Finally, the data can be transmitted to the chosen database
system. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components} provides
an architectural view of the DSP Server node with the loosely-coupled
components, highlighting the inclusion of the DSP Data Persistence.

@@

As it is depicted in figure
\ref{fig:NetBEAMS-Persistence-Server-Node-Components}, the DSP Data
Persistence component can be added into the system as an independent
plug-and-play component. In addition, the database system and the necessary
drivers are also added without any changes to the existing DSP infrastructure.
First, as DSP Messages are delivered to the server through the DSP Wire
Transport Server, as shown by flow 1, the DSP Platform is responsible to
forward a copy of the Measurement messages directly to the DSP Data
Persistence as shown by flow 2. Then, the sole responsibility of the DSP Data
Persistence is to transfer the received measurement messages to the database
system by using the connection drivers, as depicted by flow 3.

@@

The new proposed Data Persistence component must follow the DSP requirements
for a new DSP Component, which specifies the design-pattern of Data
Producer/Consumer. In a nutshell, the design of a DSP Component maintains the
dependency among 3 different major Java packages: it depends on 2 existing
packages, namely the OSGi Framework and the DSP Platform, as well as and an
additional package for the database access driver API, as shown in the UML
Package Dependency Diagram \cite{uml} of Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

The DSP Data Persistence component starts with an OSGi Activator class, and the
implementation of the DSP Component, resposible to provide services to the
platform. Based on these specifications, the following set of classes in
figure \ref{fig:DSP-DataPersistence-Activator-Class-Diagram} shows the UML
Class Diagram \cite{uml} implemented for the component.

@@

While the DSP Data Persistence Activator is responsible for the orchestration
of the classes during the execution on the OSGi Framework, the class DSP Data
Persistence extends from the DSP Platform class DSPComponent. For this reason,
this class inherits all the behavior described on \ref{netbeams-architecture}
and can be in the roles of Data Consumer and a Data Producer.

@@

The behavior of the activator class follows the specifications of the
OSGi Framework and, for this reason, instances of the DSP Data Persistence
Activator can be in the states described on the UML State Diagram depicted in
Figure \ref{fig:DSPPlatform-Install-Usage-State-Diagram}. It is important to
note that the activator is an extension of the OSGI class BundleActivator, as
well as being responsible for stopping or starting the the DSP Data Persistence
component instance. In this way, following the specifications of section
\ref{netbeams-architecture}, the DSP Data Persistence Platform can be
initialized as described as it is described by the UML Sequence Diagram of
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}:

\begin{enumerate}
  \item During the DSP Platform activation, it will get the name of the
  persistence component from the configuration artifact config.xml as seen in
  listing \ref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the DSP
  Bundle artifact is identified and installed, by creating an instance of the
  class DSP DataPersistence Activator and making a call to the method start();
  \item During the activation, an instance of the DSP Data Persistence
  component is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  "listens" the event serviceUpdate() and, and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initializes the component by calling the method initComponent();
      \item Starts the component by calling the method startComponent();
      \item Bootstrap messages are delivered to the component.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

Although the DSP Data Persistence component has been instantiated as of image
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, 
the process of delivering messages to the DSP Data Persistence has yet to be
completed. As defined by the requirements, the DSP Component must execute its
function in a concurrent fashion, using configuration parameters during its
bootstrap process. The class DSP Data Flusher is responsible for running as a
"worker thread", whose function is to verify if there are messages to be
flushed into the Database with the class Transient Persistence Layer. Figure 
\ref{fig:DSP-DataPersistence-Flusher-Classes} describes the UML Class
Diagram \cite{uml} for the main classes of this functionality.

@@

The DSP Data Flusher is a thread that is designed to be in two different
states as depicted in figure
\ref{fig:DSP-DataPersistence-Flusher-State-Diagram}: RUNNABLE and
TIMED\underline{ }WAITING. In the former state, the DSP Data Flusher will be
contacting the Transient Persistence Layer in order to verify if there are any
message in the outbound queue ready to be transmitted sent to the Database
service, while the DSP Data flusher will be waiting for the next cycle defined
by the bootstrap parameter TRANSIENT\underline{ }DATA\underline{
}FLUSHER\underline{ }DELAY in the latter state.

@@

Upon receiving a new DSP measurement message, the DSP Data Persistence
component must send its reference to the temporary memory space, as described
by the non-functional requirement in the beginning of this chapter. In this
way, the DSP Data Persistence component maintains a dependency on the Singleton
class TransientPersistenceLayer, which is responsible for keeping track of
messages that are received by working component. However, in order to add the
DSP message into its aggregated map, it first transforms the message into an
instance of the class PersistentMessageUnit, which has references to the class
regarding the location of the message, SensorLocation.

As depicted by the the UML Sequence Diagram \cite{uml} of figure
\ref{fig:From-DSPWireTransport-Server-To-DSP-Broker}, a DSP Message
is delivery to the DSP Broker for its delivery to Data Consumers configured to
receive that specific message. For this reason, Figure
\ref{fig:From-DSP-Broker-To-DSPDataPersistence-General-Sequence} continues the
steps of the DSP Broker from the gate "From DSP WireTransport Server to Broker
Gate", showing the steps taken to transfer a DSP Message to the Singleton
class TransientPersistenceLayer, responsible for maintaining a queue of
messages waiting to be sent to the database system.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. In this way,
  the DSP Data Persistence can initialize the DspDataFlusher worker thread to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY in seconds,
  among others. Listing \ref{file:dsp-data-persistence-bootstrap.xml} shows an
  implementation of a bootstrap message for the DSP Data Persistence;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. When the DSP Broker delivers a DSP MeasureMessage to this
  component, it will add the message to a temporary memory location to decrease
  I/O on the database, as depicted by the gate "create PersistentMessageUnit
  Gate". This last gate describes the simple procedure to add the message
  into the transient persistent layer. Listing
  \ref{file:dsp-message-serialized-ysi}.
\end{itemize}

The class PersistenceMessageUnit is the major transient persistence unit that
carries references regarding the originating DSP Message that was collected by
the DSP Data Persistence component. It is composed by a SensorLocation and
contains a reference to the PersistentMessageState enumaration. The former
identifies which sensor produced the collected data from the DSP Message by the
IP address, while the latter identifies whether the PersistenceMessageUnit has
been saved into the Database or not as depicted by the UML State Diagram
\cite{uml} in Figure \ref{fig:PersistentMessageState-Diagram}.

@@

According to the data model specifications defined in section
\ref{sec:dsp-persistence-data-model}, the persistence model must carry
information regarding which sensor device produced the collected data. For
this reason, an instance of the class SensorLocation carries the information
about the IP address and and optional GPS location from the component that
originally produced the DSP Message. Then, an instance of the class
PersistenceMessageUnit is created at the state ``TRANSIENT'', as detailed
on UML State Diagram \cite{uml} in Figure \ref{fig:PersistentMessageState-Diagram}.

@@

In order to build the document key and value segments as described on section
\ref{sec:dsp-persistence-data-model}. First, the value is built by extracting
the DSP Message Content from the body of the DSP Message.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.7]{../diagrams/DSP-Data-Persistence-Deployment-Sharded}
  \caption{UML Deployment Diagram for a sharded server}
  \label{fig:DSP-Data-Persistence-Deployment-Sharded}
\end{figure}

>>>>

In order to provide information to users, the RTC scientists use the
\emph{OPeNDAP} \cite{opendap} server to distribute collected data com the
\textit{SF-BEAMS} sensor network over the Internet. As discussed in the previous
chapter, the database system \emph{mongoDB} was selected for the evaluation of a
backend system for NetBEAMS. Unlike other projects reviewed in the literature, mongoDB has
never been used as a selected database for data collected from sensor devices.
In this way, the design of the proposed solution follows the requirements and
specifications detailed in chapter 4. As described before, NetBEAMS provides a
component-based infrastructure that automates the operation of SF-BEAMS sensor
network. The proposed solution presented in this chapter is based on the
NetBEAM's DSP Platform architecture, as described in chapter
\ref{chap:netbeams-overview}. The following sections describe the scenarios for
persistence, analyze the requirements captured from the RTC staff, and describe
the design and architecture of the proposed solution.

Technically speaking, this chapter includes detailed UML diagrams regarding the
different point of views of the system, following the conventions defined at
\cite{uml}.

@@

As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support SF-BEAMS, an environmental sensor network based in San
Francisco Bay, composed by different types of sensors. As shown in in Figure
\ref{fig:dsp-persistence-business-process}, the event of data collection is
started by the DSP Component deployed on each Gumstix system based on the
sampling rate frequency value. When the DSP Client component finishes the data
extraction process, a new process starts at the client side. First, each of the
co-located DSP Client processses the collected measurements and transforms
them into the alleged DSP Message, as shown in section \ref{sec:dsp-message}, 
emboding the actual collected data from the sensor devices. In this way, an 
instance of a DSP Measurement Message is added into the local temporary
outbound queue that waits to be activated. Then, as an asynchronous data
transmission event occurs within a specified rate, transferring all the queued
messages to be processed by a DSP Server host. This server host is centralized
and recives all the DSP Messages, to be forward to different messages.

@@

Once the DSP Server processes the received DSP Message, it uses the server
configuration to transmit the collected data to the Measurements
Database. As a matter of fact, the collected data from sensor devices are basic
properties with keys and associated numerical values, described in section
\ref{sec:sfbeams}. According to the descriptions of the DSP Platform and the
DSP Messages structure, the collected data from the sensors is parsed and
mapped to a POJO\footnote{Plain Old Java Object} or it can be serialized in
XML \cite{xml}. Similarly, the collected data contains attributes that
uniquely identifies them, along with provenance metadata as spatio-temporal
attributes.

@@

After analyzing the main data collection scenario and reviewing the
requirements defined by the RTC staff, this section enumerates the main
functions that the proposed solution must perform.

@@

The primary goal of system to be implemented is related to the collected data
reuse. That is, allow users to have easy access to the collected data from
\emph{NetBEAMS}. The main motivations of the proposed system is that it must
give different types of users access to the collected data without requiring
the use of SQL. Rather, it must provide a programming language abstraction
that might make the access to those data easier \cite{sn-programming-language}.
Furthermore, it must use the selected technology, mongoDB, using the
properties defined by the system such as the use of a schema-less database
systen. In this way, the designed functions for the solution, as depicted in
the UML Use Cases Diagram of
Figure \ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users}, are summarized as
follows:

@@

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: users of the
  collected data must be able to search data through the properties of the
  sensors. This is related to identity principle of what was collected
  as provenance \cite{db-provenance};
  \item \textbf{Search Measurements by Sensor Location}: users of the collected
  data must be able to search data by a given sensor location, as referred to
  the provenance from where the data was collected;
  \item \textbf{Search Measurements by Data collection date}: users of the
  collected data must be able to search data by a given date, or ranges of
  start and end date;
  \item \textbf{Annotate Existing Measurements}: users of the collected data
  must be able to annotate any existing measurements, that is, they must be
  able to add more metadata such ``tags'';
  \item \textbf{Export Data to Spreadsheets}: users of the collected data must
  be able to export the collected data to spreadsheets.
  \item \textbf{Export Data to OPEnDAP format}: users of the collected data
  must be able to export the collected data to the OPEnDAP format;
  \item \textbf{Access the data through API, Programming Languages}: users of the
  collected data must be able to search data by the use of different
  programming and scripting languages of their preference;
  \item \textbf{Remove Undesired Measurements}: users of
  the collected data must be able remove any collected data.
\end{itemize}

On the other hand, the functions performed by the solution must be designed
conformed the specifications of the Data Sensor Platform (DSP) architecture,
described in section \ref{sec:netbeams-architecture}. In this way, the
solution must be a Data Producer, implementing the following functionalities,
as shown in Figure \ref{fig:DSP-Data-Persistence-UseCases-Diagram-System}:

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-UseCases-Diagram-System}
  \caption{UML Use Case diagram for the Persistence Component}
  \label{fig:DSP-Data-Persistence-UseCases-Diagram-System}
\end{figure}

\begin{itemize}
  \item \textbf{Update Component Properties}: as described by the DSP Component
  architecture, the new DSP component must be able to have its properties
  updated. These properties must include the rate of how long it takes to the
  component flush messages;
  \item \textbf{Update Database Properties}: the component must be able to
  update the database settings such as IP Address and Port of the service;
  \item \textbf{Collect DSP Measurement Message}: as the component is designed
  to be a Data Consumer, it must be able to receive and temporarily retain the
  collected messages from the sensors;
  \item \textbf{Categorize Measurement Message}: as the DSP Message is
  delivered to the new component, it must categorize the message by its
  properties such as transaction time, originating DSP Message ID, etc;
  \item \textbf{Save Messages into the Database}: the main goal of the new
  component is to send the payload of the DSP Messages to the database in use,
  in this case, mongoDB.
\end{itemize}

@@

Different constraints were required to execute the functional requirements
described in the previous section, that is, non-functional requirements that
the system must follow:

\begin{itemize}
  \item The solution must maintain the received DSP Measurement Messages
  in-memory for a specific period of time in order to decrease the write load
  on the database (flusher rate);
  \item The Persistence storage system must be able to cope with the Data
  Volume produced by SF-BEAMS in a daily bases;
  \item The Persistence storage system must be able to scale without too much
  architectural changes, or schema changes;
  \item The Persistence storage system must be able to provide access to the
  persisted data using an abstraction similar to programming languages, as
  the authors of \cite{sn-programming-language} proposed;
  \item The Persistence storage must be able to 
\end{itemize}

@@

MongoDB is defined as a Document-Oriented Database system, that uses the
abstraction of the Key-Value-Pair data model. In this way, mongoDB does not
require the creation of a schema. However, the definition of how to describe
the collected data must be provided, as they define the keys to the system.
Similarly, in order to describe the keys, the specifications details the use of
Data Provenance as guidelines to the description of the collected data.
Therefore, the key-value pairs of any instance must include the following
properties:

@@

After analyzing the DSP Platform and the structure of the DSP Messages in
chapter \ref{sec:netbeams-architecture}, the following set of keys were designed
as be part of the data model, defined as the ``key segment'':

\begin{itemize}
  \item \textbf{message\underline{ }id}: Defines the DSP Message IP that
  collected the measurement message. This can be used to track which component generated, etc;
  \item \textbf{sensor}: Defines the properties of the sensor device;
    \begin{itemize}[label=\textbullet]
        \item \textbf{ip\underline{ }address} as the IP Address of the sensor
        device that produced the data;
        \item \textbf{location}: defines the location of the sensor device;
            \begin{itemize}[label=\textbullet]
                \item \textbf{latitude} the GPS latitude coordinate of the
                position of the sensor device, captured from references of the device;
                \item \textbf{longitude}: the GPS longitude coordinate of the
                position of the sensor device.
            \end{itemize}
    \end{itemize}
  \item \textbf{time}: identifies the time dimensions about the collected data;
    \begin{itemize}[label=\textbullet]
      \item \textbf{valid}: it is extracted from the container's date and time
      of creation, which was collected during the sensor device's reading;
      \item \textbf{transaction}: it is extracted from the DSP message
         container creation time and it is used to identify when the
         transaction occurred;
    \end{itemize} 
\end{itemize}

While the ``Key Segment'' is the set of keys formed by the Data Provenance
data, the ``Value Segment'' stores the observations from the sensor
devices. In this sensor, the observation key is defined as follows::

\begin{itemize}
  \item \textbf{observation}: this key stores the values of the document, and
  will have every different property of a given sensor device. For instance, the
  data key for the properties of the YSI device will contain the set of properties 
  defined for the device, such as battery, temperature, salinity, among other,
  as shown in Listing \ref{file:mongodb-ysi-data-format}.
\end{itemize}

@@

Considering the business analysis described in section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding Data Consumer (DC) as described
in section \ref{sec:netbeams-architecture}. Based on the Requirements
specification described earlier, this component must be responsible for
providing the processing of the DSP Measure Messages, extracting the payload of
the body of the message and converting it to the data model defined in the
previous section. Finally, the data can be transmitted to the chosen database
system. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components} provides
an architectural view of the DSP Server node with the loosely-coupled
components, highlighting the inclusion of a new DSP Component called ``DSP Data
Persistence''.

@@

As it is depicted in figure
\ref{fig:NetBEAMS-Persistence-Server-Node-Components}, the DSP Data
Persistence component can be added into the system as an independent
plug-and-play component. In addition, the database system and the necessary
drivers are also added without any changes to the existing DSP infrastructure.
In this way, a matching rule must be provided. First, as DSP Messages are
delivered to the server through the DSP Wire Transport Server, as shown by
flow 1, the DSP Platform is responsible to forward a copy of the Measurement
messages directly to the DSP Data Persistence as shown by flow 2. Then, the
sole responsibility of the DSP Data Persistence is to transfer the received
measurement messages to the database system by using the connection drivers,
as depicted by flow 3.

@@

The new proposed Data Persistence component must follow the DSP requirements
for a new DSP Component, which specifies the design-pattern of Data
Producer/Consumer. In a nutshell, the design of a DSP Component maintains the
dependency among 3 different major Java packages: it depends on 2 existing
packages, namely the OSGi Framework and the DSP Platform, as well as an
additional package for the database access driver API, as shown in the UML
Package Dependency Diagram of Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

The DSP Data Persistence component starts with an OSGi Activator class, and the
implementation of the DSP Component, resposible to provide services to the
platform. Based on these specifications, the following set of classes in
figure \ref{fig:DSP-DataPersistence-Activator-Class-Diagram} shows the UML
Class Diagram implemented for the component.

@@

While the DSP Data Persistence Activator is responsible for the orchestration
of the classes during the execution on the OSGi Framework, the class DSP Data
Persistence extends from the DSP Platform class DSPComponent. For this reason,
this class inherits all the behavior described on \ref{sec:netbeams-architecture}
and can be in the roles of Data Consumer and a Data Producer.

@@

The behavior of the activator class follows the specifications of the
OSGi Framework and, for this reason, instances of the DSP Data Persistence
Activator can be in the states described on the UML State Diagram depicted in
Figure \ref{fig:DSPPlatform-Install-Usage-State-Diagram}. It is important to
note that the activator is an extension of the OSGI class BundleActivator, as
well as being responsible for stopping or starting the the DSP Data Persistence
component instance. In this way, following the specifications of section
\ref{sec:netbeams-architecture}, the DSP Data Persistence Platform can be
initialized as described by the UML Sequence Diagram of Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}:

\begin{enumerate}
  \item During the DSP Platform activation, it will get the name of the
  persistence component from the configuration artifact config.xml as seen in
  listing \ref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the DSP
  Bundle artifact is identified and installed by creating an instance of the
  class DSP DataPersistence Activator and making a call to the method start();
  \item During the activation, an instance of the DSP Data Persistence
  component is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  "listens" the event serviceUpdate() and, and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initializes the component by calling the method initComponent();
      \item Starts the component by calling the method startComponent();
      \item Bootstrap messages are delivered to the component.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in Figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

Although the DSP Data Persistence component has been instantiated as of Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, the
process of delivering messages to the DSP Data Persistence has yet to be
completed. As defined by the requirements, the DSP Component must execute its
function in a concurrent fashion, using configuration parameters during its
bootstrap process. The class DSP Data Flusher is responsible for running as a
"worker thread", whose function is to verify if there are messages to be
flushed into the Database with the class Transient Persistence Layer. Figure
\ref{fig:DSP-DataPersistence-Flusher-Classes} describes the UML Class Diagram
for the main classes of this functionality.

@@

The DSP Data Flusher is a thread that is designed to be in two different
states as depicted in Figure
\ref{fig:DSP-DataPersistence-Flusher-State-Diagram}: RUNNABLE and
TIMED\underline{ }WAITING. In the former state, the DSP Data Flusher will be
contacting the Transient Persistence Layer in order to verify if there are any
message in the outbound queue ready to be transmitted to the Database
service, while the DSP Data flusher will be waiting for the next cycle defined
by the bootstrap parameter TRANSIENT\underline{ }DATA\underline{
}FLUSHER\underline{ }DELAY in the latter state.

@@

Upon receiving a new DSP measurement message, the DSP Data Persistence
component must send its reference to the temporary memory space, as described
by the non-functional requirement in the beginning of this chapter. In this
way, the DSP Data Persistence component maintains a dependency on the Singleton
class TransientPersistenceLayer, which is responsible for keeping track of the
messages that were received by the running component. Finally, in order to add
the DSP message into its aggregated set, it first transforms the message into an
instance of the class PersistentMessageUnit. This class carries references to
the class \emph{SensorLocation}, responsible for handling the location of the
sensor device, and consequently, the physical origin of the message.

A DSP Message is deliveried to the DSP Broker with the intent of being
forwarded to the data consumers, which are configured to receive that
specific message. This process is described by Figure
\ref{fig:From-DSPWireTransport-Server-To-DSP-Broker}. Similarly, continues the
steps of the DSP Broker from the gate "From DSP WireTransport Server to Broker
Gate", showing the steps taken to transfer a DSP Message to the Singleton
class TransientPersistenceLayer, responsible for maintaining a queue of
messages waiting to be sent to the database system.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. In this way,
  the DSP Data Persistence can initialize the DspDataFlusher worker thread to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY in seconds,
  among others. Listing \ref{file:dsp-data-persistence-bootstrap.xml} shows an
  implementation of a bootstrap message for the DSP Data Persistence;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. The component will add the message to a temporary memory location
  to decrease I/O on the database, when the DSP Broker delivers a DSP
  MeasureMessage to this component, as depicted by the gate "create
  PersistentMessageUnit Gate". This last gate describes the simple procedure
  to add the message into the transient persistent layer. Listing
  \ref{file:dsp-message-serialized-ysi}.
\end{itemize}

The class PersistenceMessageUnit is the major transient persistence unit that
carries references regarding the originating DSP Message that was collected by
the DSP Data Persistence component. It is composed by a SensorLocation and
contains a reference to the PersistentMessageState enumaration. The former
identifies which sensor produced the collected data from the DSP Message by the
IP address, while the latter identifies whether the PersistenceMessageUnit has
been saved into the Database or not as depicted by the UML State Diagram in
Figure \ref{fig:PersistentMessageState-Diagram}.

@@

According to the data model specifications defined in section
\ref{sec:dsp-persistence-data-model}, the persistence model must carry
information regarding which sensor device produced the collected data. For
this reason, an instance of the class SensorLocation carries the information
about the IP address and and optional GPS location from the component that
originally produced the DSP Message. Then, an instance of the class
PersistenceMessageUnit is created at the state ``TRANSIENT'', as detailed
on UML State Diagram in Figure \ref{fig:PersistentMessageState-Diagram}.

@@

In order to build the document key and value segments, as described on section
\ref{sec:dsp-persistence-data-model}, the reference to the DSP message is
used, as the value key is built by extracting the DSP Message Content from the
body of the DSP Message. Similarly, the key is built by gathering information
from the DSP Message and the PersistenceMessageUnit class.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.7]{../diagrams/DSP-Data-Persistence-Deployment-Sharded}
  \caption{UML Deployment Diagram for a sharded server}
  \label{fig:DSP-Data-Persistence-Deployment-Sharded}
\end{figure}

All in all, the design of the DSP Data Persistence uses the DSP Component
specification defined by the DSP Platform. First, the analysis of the business
process helped understing the requirements of the system. Similarly, the data
model design followed the specifications defined in chapter 5. In this way,
each of the layers of the system were described using UML Diagrams, having
partial details regarding the OSGi implementation. Next chapter entails the
experiments conducted to evaluate the designs discussed in this
chapter, walk-through the experiments conducted to verify the implementation
using mongoDB.
